---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# runCERIS

**Critical Environmental Regressor through Informed Search**

runCERIS identifies critical environmental windows and covariates driving
genotype-by-environment (GxE) interaction in multi-environment trials. It
performs an exhaustive search over all possible day-after-planting (DAP)
windows, correlating environmental parameters with trait means to find the
time period and covariate most predictive of GxE variation.

This R package was developed by **Dr. Yu-Ru Chen** at Iowa State University,
adapting and restructuring the original CERIS-JGRA functions created by
[Dr. Jianming Yu's lab](https://www.agron.iastate.edu/people/yu-jianming/)
at ISU. The original code is available at
[github.com/jmyu/CERIS_JGRA](https://github.com/jmyu/CERIS_JGRA). By
packaging these methods into a documented, tested R package with a Shiny
interface, runCERIS aims to make GxE environmental window analysis accessible
to the broader plant breeding community and to help users understand the
quantitative genetics concepts underlying genotype-by-environment interaction.

## Documentation

Full tutorials and vignettes: **<https://yuru0503.github.io/runCERIS/>**

## Installation

```r
# install.packages("remotes")
remotes::install_github("yuru0503/runCERIS")
```

## Quick Start

```{r quick-start, message=FALSE, warning=FALSE}
library(runCERIS)

# Load sorghum dataset
d <- load_crop_data("sorghum")

# Prepare trait data (flowering time in DAP)
exp_trait <- prepare_trait_data(d$traits, "FTdap")

# Compute environmental means and order environments
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
head(env_mean_trait)
```

```{r ceris-search, message=FALSE, warning=FALSE}
# Run CERIS search (use max_days = 80 for speed)
params <- c("DL", "GDD", "PTT", "PTR", "PTS")
ceris_result <- ceris_search(env_mean_trait, d$env_params, params, max_days = 80)

# Find the best environmental window
best <- ceris_identify_best(ceris_result, params)
cat(sprintf("Best parameter: %s (DAP %d-%d, r = %.3f)\n",
            best$param_name, best$dap_start, best$dap_end, best$correlation))
```

## Workflow Overview

| Step | Function | Description |
|------|----------|-------------|
| 1 | `load_crop_data()` | Load built-in dataset |
| 2 | `prepare_trait_data()` | Average replicates |
| 3 | `compute_env_means()` | Environmental means |
| 4 | `ceris_search()` | Exhaustive window search |
| 5 | `ceris_identify_best()` | Find best window |
| 6 | `slope_intercept()` | Reaction norm parameters |
| 7 | `jra_model()` | Joint regression |
| 8 | `cv_env()` / `cv_genotype()` | Cross-validation |
| 9 | `jgra()` / `jgra_marker()` | Genomic prediction |

## Built-in Datasets

```{r datasets}
data(crop_info)
crop_info[, c("crop", "n_envs", "traits", "has_genotype")]
```

## Shiny Application

Launch the interactive analysis interface:

```r
run_app()
```

## Vignettes

Detailed tutorials are available as package vignettes:

- [Getting Started](articles/getting-started.html)
- [Data Exploration](articles/data-exploration.html)
- [CERIS Search](articles/ceris-search.html)
- [Reaction Norms](articles/reaction-norms.html)
- [Joint Regression Analysis](articles/jra.html)
- [Cross-Validation](articles/cross-validation.html)
- [Genomic Prediction](articles/genomic-prediction.html)
- [Multi-Crop Examples](articles/multi-crop-examples.html)
- [Shiny App Guide](articles/shiny-app.html)

## References

The CERIS-JGRA methodology was developed by the scientists in
[Dr. Jianming Yu's lab](https://www.agron.iastate.edu/people/yu-jianming/)
at Iowa State University. Key publications:

- Li, X., Guo, T., Mu, Q., Li, X., & Yu, J. (2018). Genomic and
  environmental determinants and their interplay underlying phenotypic
  plasticity. *PNAS*, 115(26), 6679--6684.
  [doi:10.1073/pnas.1718326115](https://doi.org/10.1073/pnas.1718326115)

- Li, X., Guo, T., Wang, J., Bekele, W. A., Sukumaran, S., ..., & Yu, J.
  (2021). An integrated framework reinstating the environmental dimension for
  GWAS and genomic selection in crops. *Molecular Plant*, 14(6), 874--887.
  [doi:10.1016/j.molp.2021.03.010](https://doi.org/10.1016/j.molp.2021.03.010)

- Mu, Q., Guo, T., Li, X., & Yu, J. (2022). Phenotypic plasticity in plant
  height shaped by interaction between genetic loci and diurnal temperature
  range. *New Phytologist*, 233(4), 1768--1779.
  [doi:10.1111/nph.17904](https://doi.org/10.1111/nph.17904)

- Guo, T., Wei, J., Li, X., & Yu, J. (2024). Environmental context of
  phenotypic plasticity in flowering time in sorghum and rice. *Journal of
  Experimental Botany*, 75(3), 1004--1015.
  [doi:10.1093/jxb/erad398](https://doi.org/10.1093/jxb/erad398)

- Tibbs-Cortes, L. E., Guo, T., Li, X., & Yu, J. (2024). Comprehensive
  identification of genomic and environmental determinants of phenotypic
  plasticity in maize. *Genome Research*, 34(8), 1253--1265.
  [doi:10.1101/gr.279131.124](https://doi.org/10.1101/gr.279131.124)

- Wei, J., Guo, T., Mu, Q., ..., & Yu, J. (2025). Genetic and environmental
  patterns underlying phenotypic plasticity in flowering time and plant height
  in sorghum. *Plant, Cell & Environment*, 48(4), 1994--2009.
  [doi:10.1111/pce.15213](https://doi.org/10.1111/pce.15213)

## License

GPL-3
