---
title: "CERIS Search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CERIS Search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates the core CERIS analysis: an exhaustive search over
all developmental windows and environmental parameters to identify the critical
period and factor driving GxE variation. We continue with the sorghum dataset.

```{r load-data}
library(runCERIS)

sorghum <- load_crop_data("sorghum")
traits     <- sorghum$traits
env_meta   <- sorghum$env_meta
env_params <- sorghum$env_params

exp_trait      <- prepare_trait_data(traits, trait = "FTdap")
env_mean_trait <- compute_env_means(exp_trait, env_meta)

params <- c("DL", "GDD", "PTT", "PTR", "PTS")
```

## How the Exhaustive Window Search Works

CERIS tests every possible contiguous window of days after planting (DAP). A
window is defined by a start day (`Day_x`) and an end day (`Day_y`), where
`Day_x <= Day_y`. For each window and each environmental parameter, the
algorithm:

1. Computes a summary statistic (e.g., cumulative sum or mean) of the
   parameter within the window for every environment.
2. Correlates that summary with the environmental trait mean across
   environments.
3. Records the Pearson correlation (`R`) and its p-value (`P`).

This nested loop over all (start, end) pairs creates a triangular search space.
The result is a complete map of how the trait--environment association changes
with the developmental window, for every candidate parameter.

## Running the Search

`run_CERIS()` performs the exhaustive search. The `max_days` argument limits
the maximum DAP to consider, which controls both the search space and runtime.
Setting `max_days = 80` is a good starting point for sorghum flowering time:

```{r ceris-search, cache = TRUE}
ceris_result <- run_CERIS(
  env_mean_trait,
  env_params,
  params   = params,
  max_days = 80,
  loo      = FALSE,
  progress = NULL
)
```

The `loo` parameter controls whether leave-one-out cross-validation is performed
(see below). Setting `progress = NULL` suppresses the progress bar, which is
appropriate for non-interactive use.

## Understanding the Output

The result is a data frame with one row per (Day_x, Day_y) combination:

```{r result-structure}
str(ceris_result)
head(ceris_result)
```

The columns are:

| Column | Description |
|---|---|
| `Day_x` | Start DAP of the window |
| `Day_y` | End DAP of the window |
| `window` | Window length (`Day_y - Day_x + 1`) |
| `midXY` | Midpoint of the window |
| `R_<param>` | Pearson correlation for each parameter |
| `P_<param>` | P-value of the correlation for each parameter |

For example, `R_GDD` is the correlation between cumulative GDD within the window
and the trait environmental mean. High absolute correlations with low p-values
indicate that the parameter during that window is a strong predictor of GxE.

## Identifying the Best Window

`ceris_identify_best()` scans the full search result and identifies the window
and parameter with the strongest association. The `min_window` argument sets a
minimum window length to avoid spuriously narrow windows:

```{r identify-best}
best <- ceris_identify_best(ceris_result, params = params, min_window = 7)
best
```

The returned list contains the best parameter name, the start and end DAP of the
optimal window, the correlation value, and the negative log p-value. This is the
headline result of the CERIS analysis.

## Visualizing the Search: Heatmap

`plot_ceris_heatmap()` displays the full search result as a set of triangular
heatmaps, one per parameter. The x-axis is the window start day, the y-axis is
the window end day, and the color encodes the correlation strength:

```{r plot-heatmap, fig.width = 8, fig.height = 6, cache = TRUE}
plot_ceris_heatmap(ceris_result, params = params, max_days = 80)
```

Look for "hot spots" --- regions of consistently strong correlation. A well-
defined hot spot indicates a robust critical window. Diffuse patterns suggest
the association is spread across a broader developmental period.

The plot also includes marginal traces showing how the best correlation varies
with the window start and end days individually.

## Computing Window Parameters

Once you have identified the best window, use `compute_window_params()` to
calculate the parameter summary for that window in each environment. This adds a
`kPara` column to the environmental means data frame:

```{r compute-window}
env_mean_enriched <- compute_window_params(
  env_mean_trait,
  env_params,
  dap_start = best$dap_start,
  dap_end   = best$dap_end,
  params    = best$param_name
)
head(env_mean_enriched)
```

The `kPara` column contains the computed value of the best parameter within the
identified window for each environment. This is the environmental covariate that
best predicts the trait variation.

## Correlation Scatter Plot

`plot_trait_env_param()` visualizes the relationship between the trait
environmental mean and the windowed parameter:

```{r plot-scatter, fig.width = 7, fig.height = 5}
env_colors <- setNames(
  scales::hue_pal()(nrow(env_mean_trait)),
  env_mean_trait$env_code
)

plot_trait_env_param(
  env_mean_enriched,
  trait      = "FTdap",
  kpara_name = best$param_name,
  dap_start  = best$dap_start,
  dap_end    = best$dap_end,
  env_colors = env_colors
)
```

A tight linear relationship confirms that the identified window and parameter
capture the dominant environmental driver. Points that deviate from the trend
may indicate environments where additional factors are at play.

## Leave-One-Out Cross-Validation

The standard CERIS search uses all environments to compute correlations. To
assess robustness, you can enable leave-one-out (LOO) cross-validation by
setting `loo = TRUE`. In LOO mode, each environment is held out in turn and the
search is repeated on the remaining environments. This provides an honest
estimate of predictive performance:

```{r ceris-loo, cache = TRUE}
ceris_loo <- run_CERIS(
  env_mean_trait,
  env_params,
  params   = params,
  max_days = 80,
  loo      = TRUE,
  progress = NULL
)
```

The LOO result has the same structure as the standard result. You can pass it
to `ceris_identify_best()` and the visualization functions in the same way:

```{r loo-best}
best_loo <- ceris_identify_best(ceris_loo, params = params, min_window = 7)
best_loo
```

If the LOO result agrees with the full-data result (same parameter, similar
window), you can be confident that the finding is robust and not driven by a
single outlier environment.

```{r loo-heatmap, fig.width = 8, fig.height = 6, cache = TRUE}
plot_ceris_heatmap(ceris_loo, params = params, max_days = 80)
```

## Summary

The CERIS search workflow is:

1. **Run `run_CERIS()`** to compute correlations for all windows and
   parameters.
2. **Use `ceris_identify_best()`** to find the optimal window and parameter.
3. **Visualize with `plot_ceris_heatmap()`** to inspect the full search
   landscape.
4. **Extract the window parameter with `compute_window_params()`** and plot the
   trait--parameter relationship with `plot_trait_env_param()`.
5. **Validate with LOO** (`loo = TRUE`) to confirm robustness.

This systematic approach turns the complex GxE problem into an interpretable
result: a specific environmental factor, during a specific developmental window,
that explains why genotypes perform differently across environments.
