---
title: "Cross-Validation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cross-Validation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Cross-validation (CV) assesses how well a model predicts unseen data. CERIS
provides three CV strategies that differ in what is held out during prediction:

| Function | Scenario | What is dropped | Use case |
|----------|----------|-----------------|----------|
| `cv_env()` | 1-to-2 | One environment at a time | Can we predict trait values in a new environment for known genotypes? |
| `cv_genotype()` | 1-to-3 | Genotypes (K-fold) | Can we predict trait values for new genotypes in known environments? Requires marker data. |
| `cv_combined()` | 1-to-4 | Both environments and genotypes | Can we predict trait values for new genotypes in new environments? |

In addition, `loocv()` performs line-level leave-one-out cross-validation,
comparing predictions based on the environmental mean versus the CERIS-derived
kPara.

This vignette walks through all four approaches and compares their predictive
accuracy.

## Data Setup

Load the sorghum dataset and run the full CERIS pipeline to obtain the
environmental parameter (kPara):

```{r load-data}
library(runCERIS)

d <- load_crop_data("sorghum")

exp_trait <- prepare_trait_data(d$traits, "FTdap")
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
```

Run the CERIS search with `max_days = 80` for speed, then compute window
parameters:

```{r ceris-search, cache = TRUE}
ceris_result <- ceris_search(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  params         = c("DL", "GDD", "PTT", "PTR", "PTS"),
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)

best <- ceris_identify_best(ceris_result, params = c("DL", "GDD", "PTT", "PTR", "PTS"))
best

env_mean_trait <- compute_window_params(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  dap_start      = best$dap_start,
  dap_end        = best$dap_end,
  params         = best$param_name
)
```

## Line-Level Leave-One-Out CV

The `loocv()` function holds out one line at a time within each environment and
predicts its trait value using (a) the environmental mean and (b) the
CERIS-derived kPara regression. This provides a direct comparison of the two
prediction approaches:

```{r loocv}
loo_result <- loocv(exp_trait, env_mean_trait)
head(loo_result)
```

The result contains observed values (`Obs_trait`), predictions from the
environmental mean (`Prd_trait_mean`), predictions from kPara
(`Prd_trait_kPara`), and the genotype's overall mean (`Line_mean`).

Visualize the prediction accuracy with `plot_prediction_result()`:

```{r loocv-plot, fig.width = 7, fig.height = 6}
plot_prediction_result(
  obs_prd        = loo_result,
  env_mean_trait = env_mean_trait,
  trait          = "FTdap",
  kpara_name     = best$param_name
)
```

This four-panel plot compares observed versus predicted values under each
prediction method, colored by environment. Points closer to the diagonal
indicate better prediction. The kPara-based predictions typically show tighter
clustering around the diagonal when the CERIS search has identified a strong
environmental driver.

## Leave-One-Environment-Out CV

The `cv_env()` function drops one environment at a time and predicts trait
values for genotypes in that environment using the remaining environments:

```{r cv-env, cache = TRUE}
cv_env_result <- cv_env(env_mean_trait, exp_trait)
head(cv_env_result)
```

Each row contains the predicted value (`Yprd`), observed value (`Yobs`), the
held-out environment (`env_code`), and a replicate indicator (`Rep`). Compute
the overall prediction accuracy:

```{r cv-env-cor}
cor(cv_env_result$Yprd, cv_env_result$Yobs)
```

## K-Fold Genotype CV

The `cv_genotype()` function partitions genotypes into K folds, holds out each
fold in turn, and predicts trait values for the held-out genotypes using genomic
marker data. This requires:

1. A slope-intercept matrix (`lm_ab_matrix`) from `slope_intercept()` with
   kPara-based coefficients.
2. A genotype marker matrix (`SNPs`) filtered to lines present in the trial.

```{r cv-geno-prep}
lm_ab <- slope_intercept(exp_trait, env_mean_trait, type = "kPara")
SNPs <- prepare_genotype(d$genotype, unique(exp_trait$line_code))
```

Run the K-fold CV with `gFold = 5` and `gIteration = 2` for speed:

```{r cv-geno, cache = TRUE}
cv_geno_result <- cv_genotype(
  gFold          = 5,
  gIteration     = 2,
  SNPs           = SNPs,
  lm_ab_matrix   = lm_ab,
  env_mean_trait = env_mean_trait,
  exp_trait      = exp_trait,
  progress       = NULL
)
head(cv_geno_result)
```

```{r cv-geno-cor}
cor(cv_geno_result$Yprd, cv_geno_result$Yobs)
```

## Combined CV

The `cv_combined()` function simultaneously holds out both environments and
genotypes, providing the most stringent test of predictive ability:

```{r cv-combined, cache = TRUE}
cv_comb_result <- cv_combined(
  gFold          = 5,
  gIteration     = 2,
  SNPs           = SNPs,
  env_mean_trait = env_mean_trait,
  exp_trait      = exp_trait,
  progress       = NULL
)
head(cv_comb_result)
```

```{r cv-combined-cor}
cor(cv_comb_result$Yprd, cv_comb_result$Yobs)
```

## Comparing CV Methods

Use `plot_cv_results()` to display all three CV methods side by side. The
function takes a list of result data frames and a vector of labels:

```{r cv-comparison-plot, fig.width = 8, fig.height = 5}
plot_cv_results(
  cv_results = list(cv_env_result, cv_geno_result, cv_comb_result),
  labels     = c("1-to-2 (Env)", "1-to-3 (Geno)", "1-to-4 (Both)")
)
```

Each panel shows observed versus predicted values for one CV method, with
points colored by environment and the correlation coefficient displayed.
Typically, `cv_env` (dropping environments) yields the highest accuracy because
it retains all genotype information, while `cv_combined` (dropping both) is the
most challenging scenario.

## Summary

| CV Method | What is predicted | Marker data needed | Typical accuracy |
|-----------|-------------------|--------------------|------------------|
| `loocv()` | One line at a time within each environment | No | Baseline comparison |
| `cv_env()` | Known genotypes in a new environment | No | Highest |
| `cv_genotype()` | New genotypes in known environments | Yes | Moderate |
| `cv_combined()` | New genotypes in new environments | Yes | Lowest |

The gap between `cv_env` and `cv_combined` reflects how much predictive power
comes from knowing the genotype versus knowing the environment. When the gap is
small, the CERIS-derived environmental parameter captures most of the GxE
variation, making environmental characterization the dominant source of
prediction accuracy.
