---
title: "Data Exploration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Exploration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette walks through the data preparation and exploratory visualization
steps that precede a CERIS search. We use the built-in sorghum dataset
throughout.

```{r load-data}
library(runCERIS)

sorghum <- load_crop_data("sorghum")
traits     <- sorghum$traits
env_meta   <- sorghum$env_meta
env_params <- sorghum$env_params
```

## Preparing Trait Data

Raw trial data typically contains multiple replicates (lines) per environment.
`prepare_trait_data()` averages across replicates within each environment and
returns a tidy data frame with one row per line per environment:

```{r prepare-trait}
exp_trait <- prepare_trait_data(traits, trait = "FTdap")
str(exp_trait)
head(exp_trait)
```

The result has three columns: `line_code`, `env_code`, and `Yobs` (the observed
trait value). This is the starting point for all downstream analyses.

## Computing Environmental Means

`compute_env_means()` calculates the mean trait value per environment and merges
in the environment metadata. Environments are ordered by their mean, which
reveals the range of environmental effects:

```{r env-means}
env_mean_trait <- compute_env_means(exp_trait, env_meta)
str(env_mean_trait)
env_mean_trait
```

The resulting data frame includes one row per environment with `meanY` (the
environmental mean), plus all columns from `env_meta` (latitude, longitude,
planting date, etc.).

## Wide-Format Line-by-Environment Matrix

For some analyses it is useful to have a wide-format matrix with genotypes in
rows and environments in columns. `prepare_line_by_env()` creates this from the
prepared trait data and the environmental means:

```{r line-by-env}
line_env <- prepare_line_by_env(exp_trait, env_mean_trait)
head(line_env)
```

Each cell contains the observed trait value for that line in that environment.
The environments (columns) are ordered by their environmental mean.

## Visualizing Geographic Patterns

`plot_geo_order()` displays the trait distribution for each environment, ordered
by geographic or environmental-mean information. This is a useful first look at
whether GxE patterns follow a spatial gradient:

```{r plot-geo-order, fig.width = 7, fig.height = 5}
env_colors <- setNames(
  scales::hue_pal()(nrow(env_mean_trait)),
  env_mean_trait$env_code
)

plot_geo_order(exp_trait, env_mean_trait, trait = "FTdap", env_colors = env_colors)
```

Look for environments that shift in rank relative to others --- these indicate
crossover GxE interactions that CERIS is designed to dissect.

## Trait Means Across Environments

`plot_env_means()` shows individual genotype performance against the
environmental mean. Lines that cross each other indicate rank changes across
environments, the hallmark of GxE:

```{r plot-env-means, fig.width = 7, fig.height = 5}
plot_env_means(exp_trait, env_mean_trait, trait = "FTdap")
```

A fan-shaped pattern indicates scaling (variance heterogeneity) across
environments. Crossing lines indicate qualitative GxE interactions.

## Daily Environmental Factor Profiles

`plot_env_factors()` displays the daily time series of environmental parameters
for all environments. This reveals how conditions diverge across sites and
seasons:

```{r plot-env-factors, fig.width = 7, fig.height = 5}
params <- c("DL", "GDD", "PTT", "PTR", "PTS")

plot_env_factors(env_params, params = params, env_colors = env_colors)
```

Each panel shows one parameter over days after planting. Environments are
color-coded. Look for parameters that show clear separation among environments
--- these are good candidates for explaining GxE variation.

## PCA Biplot of Environments

`plot_pca_biplot()` performs PCA on windowed environmental parameters and
projects environments into a reduced space. This gives an overview of how
environments cluster based on their environmental profiles:

```{r plot-pca, fig.width = 7, fig.height = 5}
plot_pca_biplot(
  env_params,
  env_mean_trait,
  params    = params,
  group_col = "Location"
)
```

Environments that are close together in the biplot experienced similar
conditions. The parameter arrows indicate which factors drive the separation.

## Hierarchical Clustering of Environments

`plot_clustering_heatmap()` provides a complementary view by clustering
environments based on their environmental parameter profiles:

```{r plot-clustering, fig.width = 7, fig.height = 5}
plot_clustering_heatmap(env_params, params = params, scale_data = TRUE)
```

Scaling is recommended (`scale_data = TRUE`) so that all parameters contribute
equally regardless of their units. The dendrogram on the left groups similar
environments together. Compare these clusters with the geographic and
trait-based groupings to see whether environmental similarity explains GxE
patterns.

## Summary

At this stage you should have a clear picture of:

- The magnitude and pattern of GxE variation in your trait.
- Which environmental parameters vary substantially across environments.
- How environments group geographically, by trait performance, and by
  environmental profiles.

These insights guide interpretation of the CERIS search results. Proceed to
`vignette("ceris-search")` to identify the critical environmental window.
