---
title: "Data Format Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Format Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

CERIS requires four input data frames that describe the phenotypic
observations, trial environments, daily environmental covariates, and
(optionally) genotype marker data. All data frames are linked through two key
identifiers:

- **`env_code`**: A unique identifier for each trial environment (e.g.,
  location-year combination). Present in `traits`, `env_meta`, and
  `env_params`.
- **`line_code`**: A unique identifier for each genotype. Present in `traits`
  and as rownames in `genotype`.

This vignette documents the required and recommended columns for each data
frame, the derived data frames produced by the CERIS pipeline, and practical
tips for preparing custom data.

```{r load}
library(runCERIS)
d <- load_crop_data("sorghum")
```

---

## 1. traits

The `traits` data frame contains phenotypic observations at the plot level.
Each row represents one genotype observed in one environment (or one
replicate of that combination).

### Required columns

| Column | Type | Description |
|--------|------|-------------|
| `line_code` | character | Genotype identifier |
| `env_code` | character | Environment identifier |
| *(trait columns)* | numeric | One or more columns of measured trait values |

### Optional columns

| Column | Type | Description |
|--------|------|-------------|
| `pop_code` | character | Population or group identifier |
| `env_note` | character | Short description of the environment |

Any column that is not `line_code`, `env_code`, `pop_code`, or `env_note` is
treated as a trait column and must be numeric. Multiple trait columns are
allowed --- select the trait of interest when calling `prepare_trait_data()`.

### Example

```{r traits-example}
str(d$traits)
head(d$traits)
```

The sorghum dataset has two trait columns: `FTdap` (flowering time in days
after planting) and `FTgdd` (flowering time in growing degree days).

---

## 2. env_meta

The `env_meta` data frame describes each trial environment. There is one row
per environment.

### Required columns

| Column | Type | Description |
|--------|------|-------------|
| `env_code` | character | Environment identifier (must match `traits` and `env_params`) |

### Recommended columns

These columns are used by plotting functions and are strongly recommended:

| Column | Type | Description |
|--------|------|-------------|
| `lat` | numeric | Latitude of the trial site |
| `lon` | numeric | Longitude of the trial site |
| `PlantingDate` | character/Date | Planting date (e.g., "2014-05-15") |
| `TrialYear` | integer | Year of the trial |
| `Location` | character | Name of the trial location |

If `lat`, `lon`, or `PlantingDate` are missing, some visualization functions
(e.g., `plot_geo_order()`) will not work.

### Example

```{r env-meta-example}
str(d$env_meta)
d$env_meta
```

---

## 3. env_params

The `env_params` data frame contains daily environmental covariates for each
environment. Each row represents one environment on one day after planting
(DAP). This is the data CERIS searches over to identify critical
environmental windows.

### Required columns

| Column | Type | Description |
|--------|------|-------------|
| `env_code` | character | Environment identifier |
| `DAP` | integer | Day after planting (1, 2, 3, ...) |
| *(parameter columns)* | numeric | One or more daily environmental parameters |

### Available environmental parameters

The parameters used across CERIS studies are derived from three fundamental
domains --- temperature, day length, and moisture --- and their interactions
(Tibbs-Cortes et al. 2024):

| Parameter | Formula | Domain |
|-----------|---------|--------|
| DL | Day length (hours) | Day Length |
| GDD | (Tmax + Tmin) / 2 - base | Temperature |
| TMAX | Daily maximum temperature | Temperature |
| TMIN | Daily minimum temperature | Temperature |
| DTR | Tmax - Tmin | Temperature |
| PTT | GDD x DL | Temperature x Day Length |
| PTR | GDD / DL | Temperature x Day Length |
| PTD1 | (Tmax - Tmin) x DL | Temperature x Day Length |
| PTD2 | (Tmax - Tmin) / DL | Temperature x Day Length |
| PTS | (Tmax^2 - Tmin^2) x (DL)^2 | Temperature x Day Length |
| PET | Potential evapotranspiration (mm) | Temperature x Moisture |
| PRECIP | Precipitation (mm) | Moisture |
| H2O.balance | PET - PRECIP | Temperature x Moisture |

Not all parameters need to be present. You choose which parameters to search
over when calling `run_CERIS()`. The built-in datasets include:

| Crop | Parameters |
|------|------------|
| Sorghum | DL, GDD, PTT, PTR, PTS |
| Maize | TMAX, TMIN, DL, GDD, PTT, PTR, PTS |
| Rice | TMAX, TMIN, DL, GDD, PTT, PTR, PTS |
| Oat | TMAX, TMIN, DL, GDD, PTT, PTR, PTD1, PTD2, PTS |

### Example

```{r env-params-example}
str(d$env_params)
head(d$env_params)

# DAP range per environment
tapply(d$env_params$DAP, d$env_params$env_code, range)
```

---

## 4. genotype (optional)

The `genotype` matrix contains marker scores for genomic prediction. It is
optional --- CERIS search, JRA, and phenotypic cross-validation work without
it. Genomic prediction functions (`jgra()`, `jgra_marker()`, `cv_genotype()`,
`cv_combined()`) require it.

### Format

| Property | Requirement |
|----------|-------------|
| Type | Numeric matrix |
| Rows | Genotype lines (rownames must be `line_code` values) |
| Columns | Markers/SNPs (column names are marker identifiers) |
| Values | Typically coded as 0, 1, 2 (allele dosage) |

Rownames in the genotype matrix must match `line_code` values in `traits`.
Not all lines need to be present in both --- functions will use the
intersection.

### Example

```{r genotype-example}
str(d$genotype)
d$genotype[1:5, 1:5]
```

---

## 5. Derived Data Frames

The CERIS pipeline produces several intermediate data frames. Understanding
their structure helps when using downstream functions.

### exp_trait (from `prepare_trait_data()`)

Averages replicates within each line-environment combination for a single
selected trait:

```{r exp-trait}
exp_trait <- prepare_trait_data(d$traits, "FTdap")
str(exp_trait)
head(exp_trait)
```

| Column | Type | Description |
|--------|------|-------------|
| `line_code` | character | Genotype identifier |
| `env_code` | character | Environment identifier |
| `Yobs` | numeric | Mean observed trait value (averaged across replicates) |

### env_mean_trait (from `compute_env_means()`)

Environmental means merged with environment metadata:

```{r env-mean-trait}
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
str(env_mean_trait)
```

| Column | Type | Description |
|--------|------|-------------|
| `env_code` | character | Environment identifier |
| `meanY` | numeric | Mean trait value across all lines in this environment |
| *(env_meta columns)* | various | All columns from `env_meta` |

After calling `compute_window_params()`, a `kPara` column is added:

| Column | Type | Description |
|--------|------|-------------|
| `kPara` | numeric | CERIS-derived environmental parameter summarized over the best window |

### line_by_env (from `prepare_line_by_env()`)

Wide-format matrix with one row per genotype and one column per environment:

```{r line-by-env}
line_by_env <- prepare_line_by_env(exp_trait, env_mean_trait)
line_by_env[1:5, 1:4]
```

| Column | Type | Description |
|--------|------|-------------|
| `line_code` | character | Genotype identifier |
| *(env_code columns)* | numeric | Trait value for each environment |

---

## 6. Validating Your Data

Use `validate_input_data()` to check that your data meets all requirements.
The function stops with an error if a hard requirement is not met, warns about
missing recommended columns, and optionally prints a summary:

```{r validate-quiet}
# Quiet mode (default) --- returns TRUE or stops with error
validate_input_data(d$traits, d$env_meta, d$env_params, d$genotype)
```

```{r validate-verbose}
# Verbose mode --- prints a data summary
validate_input_data(d$traits, d$env_meta, d$env_params, d$genotype,
                    verbose = TRUE)
```

The genotype argument is optional:

```{r validate-no-geno}
validate_input_data(d$traits, d$env_meta, d$env_params, verbose = TRUE)
```

---

## 7. Preparing Custom Data

### Fetching weather data from NASA POWER

The easiest way to build `env_params` for your own trials is to use
`fetch_nasa_power()`. This function downloads daily temperature data from the
[NASA POWER API](https://power.larc.nasa.gov/) and automatically computes
all CERIS-derived parameters (DL, GDD, PTT, PTR, PTD1, PTD2, PTS). All you
need is the `env_meta` data frame with latitude, longitude, and planting
date for each environment:

```{r fetch-nasa-power, eval = FALSE}
# Your env_meta must have: env_code, lat, lon, PlantingDate
env_params <- fetch_nasa_power(d$env_meta, max_dap = 120, base_temp = 10)
head(env_params)

# The result is ready to use in run_CERIS()
ceris_result <- run_CERIS(
  env_mean_trait, env_params,
  params = c("DL", "GDD", "PTT", "PTR", "PTS"),
  max_days = 80
)
```

The function fetches `T2M_MAX` and `T2M_MIN` from NASA POWER, computes day
length astronomically from latitude and day of year, and derives all
photothermal parameters. No API key is required. See `?fetch_nasa_power` for
details.

### From raw weather data

If you already have daily weather records (temperature, day length), you can
compute the derived parameters and assemble the `env_params` data frame
manually. The `compile_envirome()` function can help if your data is stored
as tab-separated files on disk --- see `?compile_envirome` for details.

A minimal workflow to construct `env_params` by hand:

```{r custom-env-params, eval = FALSE}
# Suppose you have a data frame 'weather' with columns:
#   env_code, DAP, Tmax, Tmin, DL, PRECIP

weather$GDD <- pmax(0, (weather$Tmax + weather$Tmin) / 2 - 10)
weather$PTT <- weather$GDD * weather$DL
weather$PTR <- weather$GDD / weather$DL
weather$PTS <- (weather$Tmax^2 - weather$Tmin^2) * (weather$DL)^2

env_params <- weather[, c("env_code", "DAP", "DL", "GDD",
                           "PTT", "PTR", "PTS")]
```

### Common pitfalls

- **`env_code` mismatches**: Environment codes must be identical across
  `traits`, `env_meta`, and `env_params`. Check for trailing whitespace or
  inconsistent formatting (e.g., "Env1" vs "ENV1").
- **Missing DAP values**: `env_params` should have continuous DAP sequences
  within each environment (1, 2, 3, ..., N). Gaps will cause errors in the
  window search.
- **Non-numeric trait columns**: Trait columns must be numeric. If they
  contain text (e.g., "NA" as a string), convert them before analysis.
- **Genotype matrix orientation**: Rows must be lines and columns must be
  markers. If your matrix is transposed, `prepare_genotype()` will
  auto-detect and fix it.
- **Marker coding**: The genotype matrix should contain numeric values
  (typically 0, 1, 2 for biallelic markers). Factor or character columns
  will cause errors.
