---
title: "Genomic Prediction (JGRA)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Genomic Prediction (JGRA)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

Genomic prediction integrates marker data with environmental covariates to
predict genotype performance in untested environments. CERIS provides two
complementary approaches through the **JGRA** (Joint Genomic Regression
Analysis) framework:

1. **Reaction norm approach** (`jgra()`): Models genotype-by-environment
   interaction as a function of genomic breeding values and their interaction
   with an environmental covariate. The reaction norm captures how each
   genotype's performance changes along the environmental gradient identified
   by CERIS.

2. **Marker effect approach** (`jgra_marker()`): Estimates marker-specific
   sensitivity to the environmental covariate. Rather than summarizing genomic
   information into a single breeding value, this approach models individual
   marker effects and their interactions with the environment.

Both methods leverage the CERIS-identified critical environmental parameter
(kPara) to define the environmental gradient, replacing the traditional
environmental mean with a biologically interpretable covariate.

This vignette demonstrates how to:

- Prepare the required inputs for JGRA from a standard CERIS pipeline.
- Run genomic prediction using the reaction norm and marker effect approaches.
- Evaluate prediction accuracy with different cross-validation schemes.
- Forecast performance in new environments using `forecast_next_year()`.

## Data Setup

Load the sorghum dataset and run the standard CERIS pipeline through
`compute_window_params()`. This produces the kPara column needed for JGRA.

```{r load-data}
library(runCERIS)

d <- load_crop_data("sorghum")

exp_trait <- prepare_trait_data(d$traits, "FTdap")
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
```

Run the CERIS exhaustive search to identify the best environmental window:

```{r ceris-search, cache = TRUE}
params <- c("DL", "GDD", "PTT", "PTR", "PTS")

ceris_result <- run_CERIS(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  params         = params,
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)

best <- ceris_identify_best(ceris_result, params = params)
best
```

Compute the window-summarized environmental parameter for each environment:

```{r compute-kpara}
env_mean_trait <- compute_window_params(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  dap_start      = best$dap_start,
  dap_end        = best$dap_end,
  params         = best$param_name
)
head(env_mean_trait)
```

## Preparing JGRA Inputs

The `jgra()` and `jgra_marker()` functions require three primary inputs:

- **pheno**: A wide-format data frame with `line_code` in the first column
  and one column per environment containing trait values. This is the output
  of `prepare_line_by_env()`.
- **geno**: A data frame with `line_code` as the first column followed by
  marker columns. This must be constructed from the genotype matrix.
- **envir**: A data frame with `env_code` and the environmental parameter
  column to use as the covariate.

```{r prepare-inputs}
# Wide-format phenotype matrix
pheno <- prepare_line_by_env(exp_trait, env_mean_trait)
head(pheno[, 1:4])

# Genotype data frame with line_code as first column
geno_df <- data.frame(
  line_code = rownames(d$genotype),
  d$genotype,
  check.names = FALSE
)
geno_df[1:5, 1:5]

# Environmental covariate
envir <- env_mean_trait[, c("env_code", "kPara")]
envir
```

The `envir` data frame links each environment to its kPara value --- the
CERIS-identified environmental covariate summarized over the critical window.
This replaces the traditional environmental mean with a parameter that has a
direct biological interpretation (e.g., cumulative photoperiod during the
sensitive developmental period).

## Reaction Norm Approach

The `jgra()` function fits a reaction norm model that decomposes genotype
performance into a main genetic effect (breeding value) and a genotype-specific
sensitivity to the environmental covariate. The `method` argument controls the
cross-validation scheme:

- `"RM.E"`: Leave-one-environment-out --- predicts performance in environments
  excluded from training.
- `"RM.G"`: K-fold genotype cross-validation --- predicts performance of
  genotypes excluded from training.
- `"RM.GE"`: Combined --- simultaneously excludes genotypes and environments.

We start with environment leave-one-out (`RM.E`), using `fold = 5` and
`reshuffle = 2` for faster computation in this vignette:

```{r jgra-rme, cache = TRUE}
jgra_result <- jgra(
  pheno    = pheno,
  geno     = geno_df,
  envir    = envir,
  enp      = "kPara",
  env_meta = d$env_meta,
  method   = "RM.E",
  fold     = 5,
  reshuffle = 2,
  progress = NULL
)
```

The result contains prediction accuracy metrics:

```{r jgra-rme-results}
# Within-environment correlation
jgra_result$r_within

# Across-environment correlation
jgra_result$r_across

# Prediction data frame
head(jgra_result$predictions)
```

The `r_within` value measures how well the model predicts genotype rankings
within each environment, while `r_across` captures prediction accuracy across
all observations. The `predictions` data frame contains observed (`obs`) and
predicted (`pre`) values along with environment identifiers.

## Marker Effect Approach

The `jgra_marker()` function uses the same interface but estimates
marker-specific environmental sensitivities rather than summarizing genomic
information into a single breeding value. This approach can capture more
complex patterns of GxE when individual markers have differential sensitivity
to the environmental covariate:

```{r jgra-marker, eval = FALSE}
marker_result <- jgra_marker(
  pheno    = pheno,
  geno     = geno_df,
  envir    = envir,
  enp      = "kPara",
  env_meta = d$env_meta,
  method   = "RM.E",
  fold     = 5,
  reshuffle = 2,
  progress = NULL
)
```

Compare the marker effect results with the reaction norm approach:

```{r compare-methods, eval = FALSE}
comparison <- data.frame(
  Method    = c("Reaction Norm (jgra)", "Marker Effects (jgra_marker)"),
  r_within  = c(jgra_result$r_within, marker_result$r_within),
  r_across  = c(jgra_result$r_across, marker_result$r_across)
)
comparison
```

Note: `jgra_marker()` requires datasets with sufficient genotype matrix
dimensions. The sorghum example dataset is too small for the marker effect
method. With larger datasets (e.g., hundreds of lines and thousands of
markers), both methods run successfully and can be directly compared.

In many cases the two methods yield similar accuracy for environment-level
prediction (`RM.E`), but the marker effect approach may outperform the reaction
norm model when predicting untested genotypes (`RM.G` or `RM.GE`), since it
models marker-level heterogeneity in environmental sensitivity.

## Forecasting New Environments

The `forecast_next_year()` function provides a direct way to evaluate how well
the model predicts performance in environments excluded from training. This
simulates a practical breeding scenario where historical data from existing
trial sites is used to predict outcomes at new locations or in future years.

We use the first five sorghum environments as training and hold out the
remaining environments as the test set:

```{r forecast, cache = TRUE}
train_envs <- env_mean_trait$env_code[1:5]
test_envs  <- setdiff(env_mean_trait$env_code, train_envs)

cat("Training environments:", paste(train_envs, collapse = ", "), "\n")
cat("Test environments:    ", paste(test_envs, collapse = ", "), "\n")

forecast_result <- forecast_next_year(
  exp_trait      = exp_trait,
  env_mean_trait = env_mean_trait,
  trn_env        = train_envs
)
```

Visualize the forecast results with `plot_prediction_result()`, which produces
a four-panel display showing observed versus predicted values, prediction by
environment, and the relationship with kPara:

```{r plot-forecast, fig.width = 7, fig.height = 6}
plot_prediction_result(
  obs_prd        = forecast_result,
  env_mean_trait = env_mean_trait,
  trait          = "FTdap",
  kpara_name     = best$param_name
)
```

The top-left panel shows observed versus predicted trait values colored by
environment. The top-right panel displays predictions against the environmental
mean. The bottom panels show the relationship with the CERIS-identified
environmental parameter. Points close to the diagonal indicate accurate
predictions.

## Choosing the Right Method

The following table summarizes when to use each genomic prediction approach:

| Scenario | Recommended Method | Rationale |
|---|---|---|
| Predict performance in a new environment (new location or year) | `jgra()` with `method = "RM.E"` | Environment LOO directly evaluates the target scenario. Reaction norm is efficient and interpretable. |
| Predict untested genotypes in known environments | `jgra_marker()` with `method = "RM.G"` | Marker effects capture genotype-specific environmental sensitivity more accurately than a single breeding value. |
| Predict untested genotypes in new environments | `jgra_marker()` with `method = "RM.GE"` | The most challenging scenario benefits from the richer marker-level model. |
| Quick assessment with limited markers | `jgra()` with `method = "RM.E"` | The reaction norm approach is less sensitive to marker density and runs faster. |
| Large marker panels with known QTL | `jgra_marker()` | Marker-specific effects can better leverage known causal variants. |

For all methods, the prediction quality depends critically on the CERIS-derived
kPara providing a meaningful environmental gradient. A strong CERIS result
(high R-squared in the heatmap) translates directly into better genomic
prediction accuracy.

## Alternative GP Method: BayesB

Both `jgra()` and `jgra_marker()` accept a `gp_method` parameter to switch
between rrBLUP (default, ridge regression BLUP) and BayesB (Bayesian variable
selection via BGLR). BayesB assigns marker-specific shrinkage, allowing some
markers to have large effects while shrinking others to near zero --- useful
when the trait architecture involves few large-effect QTL.

To use BayesB, install the BGLR package and set `gp_method = "BayesB"`:

```{r bayesb-example, eval = FALSE}
# install.packages("BGLR")

jgra_bayesb <- jgra(
  pheno     = pheno,
  geno      = geno_df,
  envir     = envir,
  enp       = "kPara",
  env_meta  = d$env_meta,
  method    = "RM.G",
  gp_method = "BayesB",
  nIter     = 5000,
  burnIn    = 1000,
  seed      = 42,
  fold      = 5,
  reshuffle = 2
)
jgra_bayesb$r_across
```

BayesB is computationally more expensive than rrBLUP due to MCMC sampling.
For exploratory analyses, reduce `nIter` and `burnIn`. For publication-quality
results, use `nIter >= 10000` and `burnIn >= 2000`.

The `seed` parameter ensures reproducible results across runs --- this applies
to both BayesB MCMC sampling and the random fold assignments used in
cross-validation.

The same `gp_method` and `seed` parameters are available in `cv_genotype()`
and `cv_combined()`:

```{r bayesb-cv, eval = FALSE}
cv_result <- cv_genotype(
  gFold = 5, gIteration = 2,
  SNPs = SNPs, lm_ab_matrix = lm_ab,
  env_mean_trait = env_mean_trait, exp_trait = exp_trait,
  gp_method = "BayesB", nIter = 5000, burnIn = 1000, seed = 42
)
```

---

## Summary

This vignette demonstrated the two JGRA approaches for genomic prediction in
CERIS:

- **Reaction norm** (`jgra()`): Models breeding values and their interaction
  with kPara. Efficient, interpretable, and well-suited for environment-level
  prediction.
- **Marker effects** (`jgra_marker()`): Models individual marker sensitivities
  to kPara. More flexible for genotype-level prediction.
- **Forecasting** (`forecast_next_year()`): Evaluates prediction accuracy in a
  practical train/test split scenario.

Both approaches integrate the CERIS-identified critical environmental window,
connecting the environmental search results directly to genomic prediction. For
crops without genotype data, see `vignette("multi-crop-examples")` for
alternative analysis strategies.
