---
title: "Getting Started with CERIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CERIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## What is CERIS?

**CERIS** (Critical Environmental Regressor through Informed Search) is an R
package for dissecting genotype-by-environment (GxE) interactions through
exhaustive environmental window analysis. In multi-environment trials, crop
performance varies across locations and years. CERIS helps answer a fundamental
question: *which environmental factor, during which developmental window, best
explains the observed variation in a trait across environments?*

The core idea is straightforward. For every possible contiguous window of days
after planting (defined by a start day and an end day), CERIS computes a summary
of each environmental parameter (e.g., cumulative growing degree days, mean
photoperiod) and correlates that summary with the trait's environmental mean. By
exhaustively searching over all windows and all parameters, the package
identifies the critical developmental period and the environmental driver most
strongly associated with GxE variation.

## Background and Motivation

The CERIS and JGRA (Joint Genomic Regression Analysis) methods were originally
developed by **Dr. Jianming Yu's lab** at Iowa State University, with key
contributions from Xianran Li, Tingting Guo, and collaborators. The foundational
work demonstrated how environmental indices identified through exhaustive search
can be used to model phenotypic plasticity and enable genomic prediction across
environments in major crops including sorghum, maize, wheat, oat, and rice
(Li et al. 2018; Li et al. 2021; Guo et al. 2024; Wei et al. 2025).

The original code is available at
[github.com/jmyu/CERIS_JGRA](https://github.com/jmyu/CERIS_JGRA). This R
package was developed by **Yu-Ru Chen**, who adapted and restructured the
original functions into a documented, tested, and user-friendly R package. As
a researcher who frequently applies these methods, Yu-Ru Chen recognized that
packaging the code would make it far more accessible for the plant breeding
community. Beyond providing a convenient interface, the package aims to help
users understand the deeper quantitative genetics concepts behind GxE analysis
--- from reaction norms and Finlay-Wilkinson regression to genomic prediction
with environmental covariates.

## Installation

Install runCERIS from GitHub:

```r
# install.packages("remotes")
remotes::install_github("yuru0503/runCERIS")
```

Once installed, load the package:

```{r load-package}
library(runCERIS)
```

## Package Overview

CERIS provides functions organized into four categories:

| Category | Functions | Purpose |
|---|---|---|
| **Data Loading** | `load_crop_data()`, `validate_input_data()` | Load built-in datasets and validate custom data |
| **Data Preparation** | `prepare_trait_data()`, `compute_env_means()`, `prepare_line_by_env()` | Transform raw trial data into analysis-ready formats |
| **CERIS Search** | `ceris_search()`, `ceris_identify_best()`, `compute_window_params()` | Run the exhaustive window search and extract results |
| **Visualization** | `plot_ceris_heatmap()`, `plot_trait_env_param()`, `plot_geo_order()`, `plot_env_means()`, `plot_env_factors()`, `plot_pca_biplot()`, `plot_clustering_heatmap()` | Explore data and interpret search results |

## Built-in Datasets

CERIS ships with multi-environment trial data for four crops. The `crop_info`
dataset provides a quick summary:

```{r crop-info}
crop_info
```

Each crop includes between 7 and 13 environments, a set of measured traits, and
daily environmental parameters. Three crops (sorghum, maize, rice) include
genotype marker data; oat does not.

## Loading Data

Use `load_crop_data()` to load the full dataset for a crop. The function returns
a named list with four components: `traits`, `env_meta`, `env_params`, and
`genotype`.

```{r load-data}
sorghum <- load_crop_data("sorghum")
names(sorghum)
```

### Traits

The `traits` data frame contains phenotypic observations at the plot level. Each
row is one genotype in one environment:

```{r traits-str}
str(sorghum$traits)
head(sorghum$traits)
```

For sorghum, the measured traits are flowering time in days after planting
(`FTdap`) and in growing degree days (`FTgdd`). The `env_note` column provides a
short description of each environment.

### Environment Metadata

The `env_meta` data frame describes each trial environment:

```{r env-meta-str}
str(sorghum$env_meta)
sorghum$env_meta
```

This includes geographic coordinates (`lat`, `lon`), planting date, trial year,
and location name --- information used for geographic ordering and mapping.

### Environmental Parameters

The `env_params` data frame contains daily environmental covariates for each
environment. Each row is one environment on one day after planting (DAP):

```{r env-params-str}
str(sorghum$env_params)
head(sorghum$env_params)
```

For sorghum, the daily parameters are day length (`DL`), growing degree days
(`GDD`), photothermal time (`PTT`), photothermal ratio (`PTR`), and
photothermal sum (`PTS`). These are the candidate regressors that CERIS will
evaluate.

The environmental parameters used across all crops in CERIS are derived from
three fundamental domains --- temperature, day length, and moisture --- and
their interactions (Tibbs-Cortes et al. 2024):

| Parameter | Formula | Domain |
|-----------|---------|--------|
| DL | Day length (hours) | Day Length |
| GDD | (Tmax + Tmin) / 2 - base | Temperature |
| TMAX | Daily maximum temperature | Temperature |
| TMIN | Daily minimum temperature | Temperature |
| DTR | Tmax - Tmin | Temperature |
| PTT | GDD x DL | Temperature x Day Length |
| PTR | GDD / DL | Temperature x Day Length |
| PTD1 | (Tmax - Tmin) x DL | Temperature x Day Length |
| PTD2 | (Tmax - Tmin) / DL | Temperature x Day Length |
| PTS | (Tmax^2 - Tmin^2) x (DL)^2 | Temperature x Day Length |
| PET | Potential evapotranspiration (mm) | Temperature x Moisture |
| PRECIP | Precipitation (mm) | Moisture |
| H2O.balance | PET - PRECIP | Temperature x Moisture |

Not all parameters are available for every crop. The sorghum dataset includes
DL, GDD, PTT, PTR, and PTS; maize and rice add TMAX and TMIN; oat further
adds PTD1 and PTD2. See `vignette("multi-crop-examples")` for details.

### Genotype Data

The `genotype` component is a matrix of marker scores (lines in rows, markers
in columns):

```{r genotype-str}
str(sorghum$genotype)
sorghum$genotype[1:5, 1:5]
```

## Data Structure Requirements

If you bring your own data, it must follow the same structure as the built-in
datasets. The key requirements are:

- **traits**: Must contain `env_code`, `line_code`, and one or more numeric
  trait columns.
- **env_meta**: Must contain `env_code` plus location metadata. Columns `lat`,
  `lon`, and `PlantingDate` are expected by several plotting functions.
- **env_params**: Must contain `env_code`, `DAP` (days after planting), and one
  or more numeric parameter columns.
- **genotype** (optional): A numeric matrix with row names matching
  `line_code` values in `traits`.

All `env_code` values in `traits` must appear in both `env_meta` and
`env_params`.

## Validating Input Data

Use `validate_input_data()` to check that your data meets all requirements
before analysis. The function returns `TRUE` silently on success or stops with
an informative error message:

```{r validate}
validate_input_data(
  traits    = sorghum$traits,
  env_meta  = sorghum$env_meta,
  env_params = sorghum$env_params,
  genotype  = sorghum$genotype
)
```

The `genotype` argument is optional. If you do not have marker data, simply omit
it:

```{r validate-no-geno}
validate_input_data(
  traits    = sorghum$traits,
  env_meta  = sorghum$env_meta,
  env_params = sorghum$env_params
)
```

## Workflow Roadmap

A typical CERIS analysis proceeds in three stages:

1. **Data Exploration** --- Prepare trait and environment data, visualize
   geographic patterns, trait distributions, and environmental covariate
   profiles. See `vignette("data-exploration")`.

2. **CERIS Search** --- Run the exhaustive window search to identify the
   critical environmental parameter and developmental window, then visualize and
   interpret the results. See `vignette("ceris-search")`.

3. **Advanced Analysis** --- Use leave-one-out cross-validation for robustness,
   compare across crops, and integrate genotype data for genomic prediction
   (covered in the CERIS Search vignette and future extensions).

The next vignette walks through data exploration step by step.

## References

The CERIS-JGRA framework and the analyses implemented in this package are based
on the following publications from Dr. Jianming Yu's lab at Iowa State
University:

1. **Li, X.**, Guo, T., Mu, Q., Li, X., & Yu, J. (2018). Genomic and
   environmental determinants and their interplay underlying phenotypic
   plasticity. *Proceedings of the National Academy of Sciences*, 115(26),
   6679--6684.
   [doi:10.1073/pnas.1718326115](https://doi.org/10.1073/pnas.1718326115)

2. **Li, X.**, Guo, T., Wang, J., Bekele, W. A., Sukumaran, S., ..., & Yu, J.
   (2021). An integrated framework reinstating the environmental dimension for
   GWAS and genomic selection in crops. *Molecular Plant*, 14(6), 874--887.
   [doi:10.1016/j.molp.2021.03.010](https://doi.org/10.1016/j.molp.2021.03.010)

3. **Guo, T.**, Wei, J., Li, X., & Yu, J. (2024). Environmental context of
   phenotypic plasticity in flowering time in sorghum and rice. *Journal of
   Experimental Botany*, 75(3), 1004--1015.
   [doi:10.1093/jxb/erad398](https://doi.org/10.1093/jxb/erad398)

4. **Wei, J.**, Guo, T., Mu, Q., Alladassi, B. M. E., ..., & Yu, J. (2025).
   Genetic and environmental patterns underlying phenotypic plasticity in
   flowering time and plant height in sorghum. *Plant, Cell & Environment*,
   48(4), 1994--2009.
   [doi:10.1111/pce.15213](https://doi.org/10.1111/pce.15213)

5. **Tibbs-Cortes, L. E.**, Guo, T., Li, X., & Yu, J. (2024). Comprehensive
   identification of genomic and environmental determinants of phenotypic
   plasticity in maize. *Genome Research*, 34(8), 1253--1265.
   [doi:10.1101/gr.279131.124](https://doi.org/10.1101/gr.279131.124)

6. **Kusmec, A.**, Yeh, C.-T., Schnable, P. S., ..., & Yu, J. (2024).
   Data-driven identification of environmental variables influencing phenotypic
   plasticity to facilitate breeding for future climates. *New Phytologist*,
   244(2), 618--634.
   [doi:10.1111/nph.19937](https://doi.org/10.1111/nph.19937)

The original CERIS-JGRA code is available at
[github.com/jmyu/CERIS_JGRA](https://github.com/jmyu/CERIS_JGRA).
