---
title: "Statistical Foundations of Joint Regression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Foundations of Joint Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## The G$\times$E Problem

In multi-environment trials (MET), the phenotypic value $y_{ij}$ of genotype
$i$ in environment $j$ can be decomposed as:

$$y_{ij} = \mu + g_i + e_j + (ge)_{ij} + \varepsilon_{ij}$$

where $\mu$ is the grand mean, $g_i$ is the genotypic main effect, $e_j$ is
the environmental main effect, $(ge)_{ij}$ is the genotype-by-environment
interaction, and $\varepsilon_{ij}$ is the residual error.

The interaction term $(ge)_{ij}$ is the central challenge: it causes genotype
rankings to change across environments, complicating selection decisions.
Joint Regression Analysis (JRA) models this interaction as a structured,
linear function of an environmental index.

## The Finlay--Wilkinson Model

Finlay and Wilkinson (1963) proposed regressing each genotype's performance on
the environmental mean $\bar{y}_{\cdot j}$:

$$y_{ij} = \alpha_i + \beta_i \, \bar{y}_{\cdot j} + \delta_{ij}$$

where:

- $\alpha_i$ is the intercept for genotype $i$ (expected performance when
  $\bar{y}_{\cdot j} = 0$)
- $\beta_i$ is the regression coefficient (sensitivity to environmental
  quality)
- $\delta_{ij}$ is the deviation from regression (residual G$\times$E not
  explained by the linear model)

The environmental index $\bar{y}_{\cdot j} = n^{-1} \sum_{i=1}^{n} y_{ij}$
is the marginal mean of all genotypes in environment $j$, serving as a
composite measure of environmental quality.

### Adjusted intercept

Because the raw intercept $\alpha_i$ depends on the origin of the
environmental index scale, it is often more interpretable to report the
**adjusted intercept** evaluated at the grand mean:

$$\hat{\alpha}_i^{*} = \hat{\alpha}_i + \hat{\beta}_i \, \bar{\bar{y}}$$

where $\bar{\bar{y}}$ is the mean of the environmental index. This gives each
genotype's expected phenotype under average environmental conditions. In
`runCERIS`, the `Intcp_mean` column returned by `jra_model()` corresponds to
$\hat{\alpha}_i^{*}$.

## Statistical Assumptions

The validity of JRA rests on several assumptions that practitioners should
verify:

### 1. Linearity of response

The model assumes that each genotype's response to environmental quality is
**linear**. In practice, this is a first-order approximation; curvilinear
responses (e.g., optimum-type curves for temperature) will produce poor fits.
Diagnostic: inspect the per-genotype $R^2$ distribution. If a substantial
fraction of genotypes have $R^2 < 0.5$, the linear assumption may be
inadequate.

```{r linearity-check}
library(runCERIS)

d <- load_crop_data("sorghum")
exp_trait <- prepare_trait_data(d$traits, "FTdap")
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
line_by_env <- prepare_line_by_env(exp_trait, env_mean_trait)

jra_result <- jra_model(line_by_env, env_mean_trait)

# Proportion of genotypes with adequate linear fit
mean(as.numeric(jra_result$R2_mean) > 0.5)
```

### 2. Independence of the environmental index

A subtle but important issue: $\bar{y}_{\cdot j}$ is computed from the same
data used in the regression, creating a dependency between the predictor and
the response. Freeman and Perkins (1971) showed that this leads to biased
slope estimates, particularly when the number of genotypes is small ($n < 20$).
With large panels ($n > 100$), the bias is negligible because each genotype's
contribution to $\bar{y}_{\cdot j}$ is small.

### 3. Sufficient environments

The regression is fitted across environments, so statistical power depends
directly on the number of environments $J$. With fewer than 5--6 environments,
slope estimates have large standard errors and R$^2$ values are unreliable.
The sorghum dataset in `runCERIS` has 7 environments, which is a practical
minimum.

### 4. No systematic missing data

Missing combinations (genotype not observed in certain environments) are
tolerated by the regression, but if missingness is informative (e.g., a
genotype fails only in stress environments), the estimated slope will be
biased toward the mean.

## Interpreting Slopes and Intercepts

The slope $\hat{\beta}_i$ and adjusted intercept $\hat{\alpha}_i^{*}$ together
define a genotype's **adaptation profile**:

```{r adaptation-quadrants, fig.width = 5, fig.height = 4.5}
# Compute adjusted intercept at grand mean
grand_mean <- mean(env_mean_trait$meanY)

ggplot2::ggplot(
  jra_result,
  ggplot2::aes(
    x = as.numeric(.data$Slope_mean),
    y = as.numeric(.data$Intcp_mean)
  )
) +
  ggplot2::geom_point(alpha = 0.6, size = 1.5) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  ggplot2::geom_hline(
    yintercept = grand_mean,
    linetype = "dashed", color = "grey50"
  ) +
  ggplot2::annotate("text", x = 0.55, y = max(as.numeric(jra_result$Intcp_mean)),
                    label = "Stable & high\n(broad adaptation)",
                    size = 2.8, hjust = 0, color = "grey30") +
  ggplot2::annotate("text", x = 1.3, y = max(as.numeric(jra_result$Intcp_mean)),
                    label = "Responsive & high\n(favourable env.)",
                    size = 2.8, hjust = 1, color = "grey30") +
  ggplot2::annotate("text", x = 0.55, y = min(as.numeric(jra_result$Intcp_mean)),
                    label = "Stable & low\n(poor overall)",
                    size = 2.8, hjust = 0, color = "grey30") +
  ggplot2::annotate("text", x = 1.3, y = min(as.numeric(jra_result$Intcp_mean)),
                    label = "Responsive & low\n(stress-prone)",
                    size = 2.8, hjust = 1, color = "grey30") +
  ggplot2::labs(
    x = expression(paste("Slope ", hat(beta)[i])),
    y = expression(paste("Adjusted intercept ", hat(alpha)[i]^"*"))
  ) +
  ggplot2::theme_minimal(base_size = 11)
```

The four quadrants have distinct breeding implications:

- **High intercept, slope $\approx$ 1**: broadly adapted; suitable for
  diverse target environments.
- **High intercept, slope > 1**: exploits favourable environments; ideal for
  high-input systems but risky under stress.
- **Low intercept, slope < 1**: consistently poor; generally discarded.
- **Low intercept, slope > 1**: highly variable; may carry useful stress
  tolerance alleles detectable through JGRA marker analysis.

## Residual Analysis: Deviations from Regression

The deviation mean square $s_{d_i}^2$, introduced by Eberhart and Russell
(1966), quantifies how much of a genotype's G$\times$E interaction is **not**
captured by the linear model:

$$s_{d_i}^2 = \frac{1}{J - 2} \sum_{j=1}^{J} \hat{\delta}_{ij}^2$$

A genotype with high $R^2$ and low $s_{d_i}^2$ is well described by the
linear model. Conversely, a genotype with low $R^2$ exhibits non-linear
G$\times$E that warrants further investigation (e.g., via AMMI or GGE
biplot analysis).

```{r residual-analysis, fig.width = 6, fig.height = 4}
plot_jra(exp_trait, env_mean_trait, jra_result, trait = "FTdap")
```

In the right panel, the distribution of $R^2$ values across genotypes
indicates the overall adequacy of the linear model. A left-skewed distribution
(most $R^2 > 0.7$) supports JRA; a uniform or right-skewed distribution
suggests that non-linear methods may be needed.

## Limitations of the Environmental Mean

The classical JRA uses $\bar{y}_{\cdot j}$ as the sole environmental index.
This has two fundamental limitations:

1. **Circularity**: the predictor is derived from the response variable,
   inflating correlations and preventing true out-of-sample prediction.
2. **Biological opacity**: $\bar{y}_{\cdot j}$ conflates all environmental
   drivers (temperature, photoperiod, moisture, soil) into a single number,
   preventing mechanistic interpretation.

These limitations motivate the CERIS framework.

## From JRA to CERIS-Informed Regression

CERIS replaces the environmental mean with a biophysically meaningful
covariate $h_j$ aggregated over a critical developmental window:

$$y_{ij} = \alpha_i + \beta_i \, h_j + \delta_{ij}$$

where $h_j$ is (for example) mean photoperiod between DAP 20 and 55 for
environment $j$. This substitution has several advantages:

- **Predictive power**: $h_j$ can be computed for new environments using only
  meteorological data, enabling true forecasting without observed phenotypes.
- **Biological interpretation**: the slope $\beta_i$ now quantifies the
  genotype's sensitivity to a specific environmental factor during a specific
  developmental stage.
- **Higher $R^2$**: when the correct factor and window are identified, the
  covariate-based regression typically explains more G$\times$E variance than
  the environmental mean.

```{r ceris-vs-mean}
params <- c("DL", "GDD", "PTT", "PTR", "PTS")
ceris_result <- ceris_search(env_mean_trait, d$env_params, params, max_days = 80)
best <- ceris_identify_best(ceris_result, params = params)

env_mean_trait <- compute_window_params(
  env_mean_trait, d$env_params,
  best$dap_start, best$dap_end, best$param_name
)

si_both <- slope_intercept(exp_trait, env_mean_trait, type = "both")

# Compare R-squared: environmental mean vs CERIS parameter
r2_comparison <- data.frame(
  Predictor = c("Environmental mean", "CERIS kPara"),
  Median_R2 = c(
    median(as.numeric(si_both$R2_para), na.rm = TRUE),
    median(as.numeric(si_both$R2_para), na.rm = TRUE)
  )
)
```

```{r ceris-improvement}
# Median R2 using kPara
median(as.numeric(si_both$R2_para), na.rm = TRUE)
```

## Practical Workflow

A recommended workflow for JRA within the `runCERIS` framework:

1. **Fit classical JRA** with `jra_model()` to establish a baseline.
2. **Inspect $R^2$ distribution** --- if median $R^2 > 0.7$, the linear model
   is adequate; otherwise consider non-linear alternatives.
3. **Run CERIS search** with `ceris_search()` to identify the best
   environmental parameter and developmental window.
4. **Re-fit with kPara** using `slope_intercept(type = "both")` to compare
   the covariate-based regression against the mean-based baseline.
5. **Interpret slopes** biologically in terms of the identified environmental
   factor (e.g., photoperiod sensitivity during vegetative phase).
6. **Validate** with leave-one-environment-out cross-validation via `loocv()`
   to confirm that the improved fit translates to predictive accuracy.

## References

- Finlay, K.W. and Wilkinson, G.N. (1963). The analysis of adaptation in a
  plant-breeding programme. *Australian Journal of Agricultural Research*,
  **14**(6), 742--754.
- Eberhart, S.A. and Russell, W.A. (1966). Stability parameters for comparing
  varieties. *Crop Science*, **6**(1), 36--40.
- Freeman, G.H. and Perkins, J.M. (1971). Environmental and
  genotype-environmental components of variability. VIII. Relations between
  genotypes grown in different environments and measures of these environments.
  *Heredity*, **27**(1), 15--23.
- Li, X. and Guo, T. et al. CERIS: Critical Environmental Regressor through
  Informed Search for dissecting genotype-by-environment interaction.
