---
title: "Joint Regression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Joint Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

**Joint Regression Analysis (JRA)** is a classical method for characterizing
genotype-by-environment (GxE) interaction. Originally proposed by Finlay and
Wilkinson (1963) and refined by Eberhart and Russell (1966), JRA regresses each
genotype's performance on the environmental mean. The resulting slope and
intercept summarize how each genotype responds to environmental quality.

CERIS implements JRA through the `jra_model()` function. This vignette
demonstrates how to fit the model, visualize the results, and interpret the
output.

## Data Setup

Load the sorghum dataset and prepare the inputs required by `jra_model()`:

```{r load-data}
library(runCERIS)

d <- load_crop_data("sorghum")

exp_trait <- prepare_trait_data(d$traits, "FTdap")
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
line_by_env <- prepare_line_by_env(exp_trait, env_mean_trait)
```

The `line_by_env` matrix is a wide-format table with genotypes in rows and
environments in columns, containing the observed trait values. This is the
primary input to the JRA model.

```{r line-by-env-preview}
line_by_env[1:5, 1:5]
```

## Fitting the JRA Model

The `jra_model()` function fits a simple linear regression for each genotype
against the environmental means:

```{r jra-fit}
jra_result <- jra_model(line_by_env, env_mean_trait)
head(jra_result)
```

The output data frame contains one row per genotype with the following columns:

| Column | Description |
|--------|-------------|
| `line_code` | Genotype identifier |
| `Intcp` | Raw intercept from the regression |
| `Intcp_mean` | Intercept adjusted to the mean environmental index |
| `Slope_mean` | Slope of the regression on the environmental mean |
| `R2_mean` | Coefficient of determination |

## Summary Statistics

A quick summary of the slope and R-squared distributions reveals the overall
pattern of GxE interaction in the trial:

```{r jra-summary}
summary(jra_result$Slope_mean)
summary(jra_result$R2_mean)
```

The mean slope across all genotypes should be close to 1.0 (by construction,
since the environmental index is the mean across genotypes). The spread of
slopes indicates how much genotypes differ in their responsiveness.

## Visualization

The `plot_jra()` function produces a two-panel display:

- **Left panel**: Regression lines for all genotypes plotted against the
  environmental mean, with observed data points.
- **Right panel**: Histogram of R-squared values across genotypes.

```{r jra-plot, fig.width = 7, fig.height = 5}
plot_jra(exp_trait, env_mean_trait, jra_result, trait = "FTdap")
```

In the left panel, the spread of regression lines shows the degree of GxE
interaction. Parallel lines would indicate no crossover interaction (genotype
rankings remain constant), while crossing lines indicate rank changes across
environments.

The R-squared histogram shows how well the simple linear model fits each
genotype. High R-squared values across most genotypes suggest that a linear
model captures the main pattern of GxE interaction. Genotypes with low
R-squared may exhibit non-linear responses or be influenced by factors not
captured by the environmental mean.

## Interpreting Slopes

The slope from JRA has a direct biological interpretation:

| Slope | Interpretation |
|-------|----------------|
| > 1 | **Responsive**: The genotype is above average in good environments and below average in poor ones. It exploits favorable conditions but is vulnerable to stress. |
| ~ 1 | **Average**: The genotype tracks the environmental mean proportionally. Its performance mirrors the trial average. |
| < 1 | **Stable**: The genotype is less affected by environmental variation. It performs relatively better in poor environments but does not fully capitalize on good ones. |

Breeders targeting broad adaptation prefer genotypes with high intercepts
(good overall performance) and slopes near 1. Those targeting specific
environments may select responsive genotypes (slope > 1) for favorable
conditions or stable genotypes (slope < 1) for stress-prone regions.

## Relationship to CERIS

JRA uses the environmental mean as the environmental index, which is a simple
and widely used approach. However, the environmental mean conflates all sources
of environmental variation into a single number. CERIS improves on this by
identifying a specific environmental parameter and developmental window that
best explains GxE variation. The slope-intercept analysis in
`vignette("reaction-norms")` extends JRA by using the CERIS-derived kPara as
the environmental index, often yielding higher R-squared values and more
biologically interpretable results.
