---
title: "Multi-Crop Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Crop Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

CERIS ships with multi-environment trial data for four crops: sorghum, maize,
rice, and oat. Each crop differs in the number of environments, available
traits, environmental parameters, and whether genotype marker data is included.
This vignette demonstrates the full CERIS workflow for maize, rice, and oat,
highlighting how to adapt the analysis to each crop's characteristics.

The sorghum workflow is covered in detail in earlier vignettes
(`vignette("getting-started")` through `vignette("genomic-prediction")`).
Here we focus on the remaining three crops.

## Crop Overview

The `crop_info` dataset provides a quick summary of all available crops:

```{r crop-info}
library(runCERIS)

crop_info
```

Key differences across crops:

| Crop | Environments | Traits | Env Parameters | Genotype Data |
|---|---|---|---|---|
| Sorghum | 7 | FTdap, FTgdd | DL, GDD, PTT, PTR, PTS | Yes |
| Maize | 10 | FT, PH | TMAX, TMIN, DL, GDD, PTT, PTR, PTS | Yes |
| Rice | 9 | FTdap, FTgdd | TMAX, TMIN, DL, GDD, PTT, PTR, PTS | Yes |
| Oat | 13 | FTdap, PH, GY | TMAX, TMIN, DL, GDD, PTT, PTR, PTD1, PTD2, PTS | No |

Oat is notable for having no genotype data, which limits the analysis to
phenotypic methods. Oat also includes two additional environmental parameters
(`PTD1` and `PTD2`) not present in the other crops.

---

## Maize: Full Workflow

Maize provides 10 environments, two traits (flowering time `FT` and plant
height `PH`), seven environmental parameters, and genotype marker data. We
demonstrate the complete pipeline using flowering time.

### Data Loading and Preparation

```{r maize-load}
maize <- load_crop_data("maize")
names(maize)

str(maize$traits)
str(maize$env_params)
```

Prepare the trait data and compute environmental means:

```{r maize-prepare}
maize_exp <- prepare_trait_data(maize$traits, "FT")
maize_env <- compute_env_means(maize_exp, maize$env_meta)

head(maize_exp)
maize_env
```

### CERIS Search

Run the exhaustive window search over all seven environmental parameters. With
10 environments, the search is somewhat more computationally intensive than
sorghum's 7 environments:

```{r maize-search, cache = TRUE}
maize_params <- c("TMAX", "TMIN", "DL", "GDD", "PTT", "PTR", "PTS")

maize_ceris <- ceris_search(
  env_mean_trait = maize_env,
  env_params     = maize$env_params,
  params         = maize_params,
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)
```

### Identifying the Best Window

```{r maize-best}
maize_best <- ceris_identify_best(maize_ceris, params = maize_params)
maize_best
```

### Heatmap Visualization

The heatmap reveals the correlation landscape across all window-parameter
combinations. Look for a well-defined peak in the heatmap --- this indicates a
clear critical window.

```{r maize-heatmap, fig.width = 8, fig.height = 6}
plot_ceris_heatmap(
  ceris_result = maize_ceris,
  params       = maize_params,
  max_days     = 80
)
```

### Computing Window Parameters

```{r maize-kpara}
maize_env <- compute_window_params(
  env_mean_trait = maize_env,
  env_params     = maize$env_params,
  dap_start      = maize_best$dap_start,
  dap_end        = maize_best$dap_end,
  params         = maize_best$param_name
)
head(maize_env)
```

### Joint Regression Analysis

The JRA model regresses genotype performance on the environmental mean and the
CERIS-derived kPara. This quantifies how much of the GxE variation is explained
by the identified environmental window:

```{r maize-jra}
maize_line_env <- prepare_line_by_env(maize_exp, maize_env)

maize_jra <- jra_model(maize_line_env, maize_env)
maize_jra
```

### Cross-Validation

Environment-level cross-validation evaluates prediction accuracy when one
environment is held out at a time:

```{r maize-cv}
maize_cv <- cv_env(maize_env, maize_exp)
maize_cv
```

This gives a realistic estimate of how well the model would predict performance
in an untested environment, which is the primary use case in breeding programs.

---

## Rice: Abbreviated Workflow

Rice provides 9 environments, two traits (`FTdap` and `FTgdd`), and the same
seven environmental parameters as maize. We demonstrate a streamlined workflow
focusing on the CERIS search and visualization.

### Data Preparation

```{r rice-load}
rice <- load_crop_data("rice")

rice_exp <- prepare_trait_data(rice$traits, "FTdap")
rice_env <- compute_env_means(rice_exp, rice$env_meta)
rice_env
```

### CERIS Search

```{r rice-search, cache = TRUE}
rice_params <- c("TMAX", "TMIN", "DL", "GDD", "PTT", "PTR", "PTS")

rice_ceris <- ceris_search(
  env_mean_trait = rice_env,
  env_params     = rice$env_params,
  params         = rice_params,
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)
```

### Best Window and Heatmap

```{r rice-best}
rice_best <- ceris_identify_best(rice_ceris, params = rice_params)
rice_best
```

```{r rice-heatmap, fig.width = 8, fig.height = 6}
plot_ceris_heatmap(
  ceris_result = rice_ceris,
  params       = rice_params,
  max_days     = 80
)
```

Rice is a short-day plant, so day length (`DL`) and photoperiod-derived
parameters (`PTT`, `PTR`, `PTS`) are often the strongest predictors of
flowering time variation across environments. Compare the heatmap pattern with
maize to see how the critical window and parameter differ between crops.

### Computing Window Parameters

```{r rice-kpara}
rice_env <- compute_window_params(
  env_mean_trait = rice_env,
  env_params     = rice$env_params,
  dap_start      = rice_best$dap_start,
  dap_end        = rice_best$dap_end,
  params         = rice_best$param_name
)
head(rice_env)
```

Rice has genotype marker data available, so the full JGRA pipeline from
`vignette("genomic-prediction")` can be applied. The workflow is identical
to sorghum --- prepare the wide-format phenotype matrix, construct the genotype
data frame, and call `jgra()` or `jgra_marker()`.

---

## Oat: Analysis Without Genotype Data

Oat is the largest dataset with 13 environments, three traits (`FTdap`, `PH`,
`GY`), and nine environmental parameters including the unique `PTD1` and `PTD2`
parameters. Crucially, oat has **no genotype marker data**, which restricts the
analysis to phenotypic methods.

### Data Loading

```{r oat-load}
oat <- load_crop_data("oat")
names(oat)

# Confirm no genotype data
is.null(oat$genotype)

str(oat$traits)
str(oat$env_params)
```

### Data Preparation

We use flowering time in days after planting (`FTdap`) as the trait of
interest:

```{r oat-prepare}
oat_exp <- prepare_trait_data(oat$traits, "FTdap")
oat_env <- compute_env_means(oat_exp, oat$env_meta)
oat_env
```

### CERIS Search with Extended Parameters

Oat includes nine environmental parameters --- two more than the other crops.
The additional parameters `PTD1` and `PTD2` represent alternative
photothermal-day formulations that may capture different aspects of the
photoperiod-temperature interaction:

```{r oat-search, cache = TRUE}
oat_params <- c("TMAX", "TMIN", "DL", "GDD", "PTT", "PTR", "PTD1", "PTD2", "PTS")

oat_ceris <- ceris_search(
  env_mean_trait = oat_env,
  env_params     = oat$env_params,
  params         = oat_params,
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)
```

### Best Window and Visualization

```{r oat-best}
oat_best <- ceris_identify_best(oat_ceris, params = oat_params)
oat_best
```

```{r oat-heatmap, fig.width = 8, fig.height = 6}
plot_ceris_heatmap(
  ceris_result = oat_ceris,
  params       = oat_params,
  max_days     = 80
)
```

With 13 environments, oat provides the most statistical power for the CERIS
search among the four built-in datasets. The heatmap may reveal multiple
competing windows or parameters, particularly among the correlated
photoperiod-derived parameters. The `ceris_identify_best()` function selects
the single best combination, but inspecting the heatmap visually can reveal
near-optimal alternatives worth investigating.

### Cross-Validation (Phenotypic Only)

Since oat has no genotype data, genomic prediction methods (`jgra()`,
`jgra_marker()`) cannot be applied. The analysis is restricted to phenotypic
cross-validation:

```{r oat-kpara}
oat_env <- compute_window_params(
  env_mean_trait = oat_env,
  env_params     = oat$env_params,
  dap_start      = oat_best$dap_start,
  dap_end        = oat_best$dap_end,
  params         = oat_best$param_name
)
```

```{r oat-cv}
oat_cv <- cv_env(oat_env, oat_exp)
oat_cv
```

Environment-level cross-validation remains fully functional without marker
data. This evaluates how well the CERIS-identified environmental parameter
predicts trait variation in held-out environments based on phenotypic data
alone.

For oat, you can also explore the other available traits (`PH` for plant
height, `GY` for grain yield) by repeating the pipeline with a different
`trait` argument in `prepare_trait_data()`. Different traits may have different
critical windows and environmental drivers.

---

## Cross-Crop Comparison

The four crops illustrate the flexibility of the CERIS framework:

| Aspect | Sorghum | Maize | Rice | Oat |
|---|---|---|---|---|
| Best parameter | Varies by run | Varies by run | Varies by run | Varies by run |
| Photoperiod sensitive | Yes (short-day) | Less so | Yes (short-day) | Yes (long-day) |
| Genomic prediction | Full JGRA | Full JGRA | Full JGRA | Not available |
| Number of parameters | 5 | 7 | 7 | 9 |
| Unique features | Baseline dataset | Temperature params | Temperature params | PTD1, PTD2; no genotype |

Key observations when comparing across crops:

1. **Parameter set size matters.** More parameters mean a larger search space.
   The oat analysis evaluates 9 parameters compared to sorghum's 5, which
   increases the risk of spurious associations but also provides a richer
   characterization of the environment.

2. **Photoperiod sensitivity varies.** Sorghum and rice are short-day plants
   where photoperiod-derived parameters often dominate. Maize flowering is less
   photoperiod-sensitive, so temperature-related parameters may emerge as
   stronger predictors. Oat is a long-day plant with yet another pattern.

3. **Genotype data enables prediction.** The three crops with marker data
   support the full JGRA pipeline for genomic prediction. Without markers
   (oat), the analysis focuses on understanding GxE patterns through
   environmental characterization and phenotypic cross-validation.

## Summary

This vignette demonstrated CERIS workflows for three crops beyond sorghum:

- **Maize**: Full pipeline including JRA and cross-validation, with seven
  environmental parameters and genotype data.
- **Rice**: Streamlined search and visualization, with the same parameter set
  as maize and potential for full JGRA analysis.
- **Oat**: Extended parameter set (nine parameters including PTD1 and PTD2)
  but no genotype data, restricting analysis to phenotypic methods.

The CERIS search framework is crop-agnostic --- the same functions work across
all crops. The key decisions are which traits and parameters to include, and
whether genotype data is available for genomic prediction. Use `crop_info` to
quickly assess what is available for each built-in dataset.
