---
title: "Reaction Norms and Slope-Intercept Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reaction Norms and Slope-Intercept Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

A **reaction norm** describes how a genotype's phenotype changes across
environments. In the context of multi-environment trials, reaction norms
visualize each genotype's response as a line plotted against an environmental
index. The classical framework is the Finlay-Wilkinson regression, where
individual genotype performance is regressed on the environmental mean or on a
CERIS-derived environmental parameter (kPara).

This vignette demonstrates how to:

1. Compute and visualize reaction norms using both environmental means and
   CERIS-identified parameters.
2. Extract slope-intercept statistics that characterize genotype stability and
   responsiveness.
3. Interpret pairwise distance plots for genotype comparison.

## Data Setup

Load the sorghum dataset and prepare the trait data through the full CERIS
pipeline. We use flowering time in days after planting (`FTdap`) as the trait
of interest.

```{r load-data}
library(runCERIS)

d <- load_crop_data("sorghum")

exp_trait <- prepare_trait_data(d$traits, "FTdap")
head(exp_trait)
```

Compute environmental means by merging trait averages with environment metadata:

```{r env-means}
env_mean_trait <- compute_env_means(exp_trait, d$env_meta)
head(env_mean_trait)
```

Run the CERIS exhaustive search to identify the critical environmental window.
We set `max_days = 80` to keep computation manageable:

```{r ceris-search, cache = TRUE}
ceris_result <- ceris_search(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  params         = c("DL", "GDD", "PTT", "PTR", "PTS"),
  max_days       = 80,
  loo            = FALSE,
  progress       = NULL
)
```

Identify the best window and compute the corresponding environmental parameter
(kPara) for each environment:

```{r best-window}
best <- ceris_identify_best(ceris_result, params = c("DL", "GDD", "PTT", "PTR", "PTS"))
best

env_mean_trait <- compute_window_params(
  env_mean_trait = env_mean_trait,
  env_params     = d$env_params,
  dap_start      = best$dap_start,
  dap_end        = best$dap_end,
  params         = best$param_name
)
head(env_mean_trait)
```

The `env_mean_trait` data frame now contains a `kPara` column --- the
CERIS-identified environmental covariate summarized over the best window.

## Reaction Norm Plot

The `plot_reaction_norm()` function produces a four-panel display:

1. **Top-left**: Reaction norms plotted against the environmental mean (meanY).
2. **Top-right**: Reaction norms plotted against kPara.
3. **Bottom-left**: Distribution of environmental means.
4. **Bottom-right**: Distribution of kPara values.

```{r reaction-norm-plot, fig.width = 7, fig.height = 5}
plot_reaction_norm(exp_trait, env_mean_trait, trait = "FTdap")
```

The top panels show each genotype as a line connecting its observed values
across environments. Steeper lines indicate genotypes that are more responsive
to environmental variation, while flatter lines indicate more stable genotypes.
Comparing the two top panels reveals whether the CERIS-derived kPara provides a
cleaner separation of environments than the simple mean.

## Slope-Intercept Analysis

The `slope_intercept()` function regresses each genotype's observations on
either the environmental mean, the CERIS parameter (kPara), or both. This
yields a slope (responsiveness) and intercept (baseline performance) for each
genotype.

### Using kPara

```{r slope-kpara}
res_para <- slope_intercept(exp_trait, env_mean_trait, type = "kPara")
head(res_para)
```

The output contains:

- `Intcp_para_adj`: Intercept adjusted to the mean kPara value.
- `Intcp_para`: Raw intercept from the regression on kPara.
- `Slope_para`: Slope of the regression on kPara.
- `R2_para`: Coefficient of determination for the kPara regression.

### Using Environmental Mean

```{r slope-mean}
res_mean <- slope_intercept(exp_trait, env_mean_trait, type = "mean")
head(res_mean)
```

This returns `Intcp_mean` and `Slope_mean` --- the classical Finlay-Wilkinson
regression coefficients.

### Both Together

```{r slope-both}
res_both <- slope_intercept(exp_trait, env_mean_trait, type = "both")
head(res_both)
```

Using `type = "both"` returns all columns from both regressions in a single
data frame, which is convenient for comparing the two approaches.

## Slope-Intercept Plot

The `plot_slope_intercept()` function produces a two-panel display showing the
relationship between intercept (baseline performance) and slope
(responsiveness). The first argument must be the trait data merged with the
kPara values from `env_mean_trait`:

```{r slope-intercept-plot, fig.width = 7, fig.height = 5}
exp_trait_merged <- merge(
  exp_trait,
  env_mean_trait[, c("env_code", "kPara")],
  by = "env_code"
)

plot_slope_intercept(
  exp_trait_merged = exp_trait_merged,
  res_para         = res_para,
  trait            = "FTdap",
  kpara_name       = best$param_name
)
```

The left panel shows regression lines for each genotype against kPara, while
the right panel displays the slope versus adjusted intercept. Genotypes in
different quadrants have distinct adaptation profiles.

## Pairwise Distance Plot

The `plot_pairwise_dist()` function produces a four-panel display that
characterizes the pairwise distances between environments and between genotypes:

```{r pairwise-dist-plot, fig.width = 7, fig.height = 5}
plot_pairwise_dist(exp_trait, env_mean_trait, trait = "FTdap")
```

These panels help identify clusters of similar environments or genotypes and
highlight outliers that behave differently from the rest.

## Interpreting Stability and Responsiveness

The slope from the Finlay-Wilkinson regression (or its CERIS-enhanced
counterpart) is the primary indicator of genotype adaptation:

| Slope | Interpretation |
|-------|----------------|
| > 1 | **Responsive**: performs disproportionately well in favorable environments but poorly in unfavorable ones. |
| ~ 1 | **Average**: tracks the environmental mean proportionally. |
| < 1 | **Stable**: less sensitive to environmental variation; maintains more consistent performance. |

A high R-squared value indicates that the environmental index (mean or kPara)
explains most of the genotype's variation across environments. Low R-squared
suggests that other factors (e.g., biotic stress, soil heterogeneity) are
driving the genotype's response.

When comparing the mean-based and kPara-based regressions, an improvement in
R-squared with kPara indicates that the CERIS-identified environmental window
captures GxE variation more effectively than the overall environmental mean.

```{r compare-r2}
summary(res_both$R2_para)
```

Genotypes with high R-squared under kPara but low R-squared under the
environmental mean are those whose performance is specifically driven by the
critical environmental factor identified by CERIS.
