---
title: "Shiny Application Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shiny Application Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

CERIS includes an interactive Shiny application that provides a point-and-click
interface for the entire analysis pipeline. The app is designed for researchers
who prefer a graphical workflow or who want to quickly explore results before
writing analysis scripts. Every analysis available in the app can also be
performed programmatically using the functions documented in the other
vignettes.

## Launching the App

Start the Shiny application with a single function call:

```{r launch}
library(runCERIS)

run_app()
```

The app opens in your default web browser. All computation runs locally in your
R session.

## Tab-by-Tab Walkthrough

The application is organized into eight tabs, each corresponding to a stage of
the CERIS analysis workflow.

### 1. Data Tab

The Data tab is the entry point for every analysis. You can either select a
built-in crop dataset or upload your own files.

**Using built-in data:**

Select a crop from the dropdown menu (sorghum, maize, rice, or oat). The data
loads immediately and a summary of the dataset is displayed, including the
number of environments, lines, traits, and available environmental parameters.

**Uploading custom data:**

Upload four CSV files corresponding to the four data components. Each file must
follow the column structure described below (see [Custom Data Format
Requirements](#custom-data-format-requirements)). After uploading, the app
validates the data and reports any formatting issues.

Once data is loaded, select the trait of interest from the trait dropdown. This
selection propagates to all downstream tabs.

### 2. Explore Tab

The Explore tab provides interactive visualizations for understanding the data
before running the CERIS search:

```{r explore-example}
# These visualizations are available in the Explore tab:
# - Geographic ordering of environments
# - Trait means across environments (Finlay-Wilkinson style)
# - Daily environmental factor profiles
# - PCA biplot of environments
# - Hierarchical clustering heatmap
```

- **Geographic Order**: Displays boxplots of the selected trait for each
  environment, ordered by latitude or environmental mean. This reveals spatial
  gradients in trait performance.

- **Environment Means**: Shows individual genotype performance lines plotted
  against the environmental mean. Crossing lines indicate GxE interactions.

- **Environmental Factors**: Time-series plots of daily environmental
  parameters across all environments. Use these to identify parameters that
  show strong separation among environments.

- **PCA Biplot**: Projects environments into a reduced space based on their
  environmental parameter profiles. Parameter arrows indicate which factors
  drive the separation.

- **Clustering Heatmap**: Groups environments by similarity in their
  environmental profiles. Compare these clusters with the trait-based groupings
  to assess whether environmental similarity explains GxE patterns.

All plots can be customized with color palettes and exported as PNG or PDF
files.

### 3. CERIS Tab

The CERIS tab runs the core exhaustive window search:

```{r ceris-example}
# In the CERIS tab:
# 1. Select environmental parameters to include in the search
# 2. Set max_days (maximum window endpoint)
# 3. Optionally enable leave-one-out mode
# 4. Click "Run Search"
# 5. View the heatmap and best window results
```

- **Parameter Selection**: Check the parameters to include in the search. All
  available parameters are listed based on the loaded dataset.

- **Max Days**: Set the maximum number of days after planting to consider. A
  value of 80 is a reasonable default for most crops.

- **Leave-One-Out**: When enabled, the search performs leave-one-environment-out
  cross-validation at each window, which provides a more robust but slower
  assessment of window quality.

- **Run Search**: Initiates the exhaustive search. A progress indicator shows
  the current parameter being evaluated. Computation time depends on the number
  of parameters, environments, and the max_days setting.

- **Heatmap**: After the search completes, an interactive heatmap displays the
  correlation (R-squared) between the windowed environmental parameter and the
  trait across all start-day and end-day combinations. The best window is
  highlighted.

- **Best Window**: A summary panel shows the identified best parameter, window
  start and end days, and the corresponding R-squared value. Click "Apply" to
  pass these results to downstream tabs.

### 4. Reaction Norm Tab

The Reaction Norm tab visualizes how genotype performance changes across the
environmental gradient:

```{r rn-example}
# In the Reaction Norm tab:
# - View reaction norms against environmental mean and kPara
# - Explore slope-intercept analysis
# - Identify stable vs. responsive genotypes
```

- **Reaction Norm Plots**: Four-panel display showing genotype-level reaction
  norms against both the environmental mean and the CERIS-identified kPara.
  Lines connecting a genotype's performance across environments reveal its
  sensitivity pattern.

- **Slope-Intercept Analysis**: Scatter plot of genotype slopes (environmental
  sensitivity) versus intercepts (baseline performance). Genotypes are
  categorized as stable (slope < 1), average (slope near 1), or responsive
  (slope > 1).

- **Pairwise Distances**: Characterizes the similarity structure among
  environments and among genotypes based on their trait values.

### 5. JRA Tab

The JRA (Joint Regression Analysis) tab fits the classical joint regression
model:

```{r jra-example}
# In the JRA tab:
# - View joint regression results
# - Compare R-squared for mean-based vs. kPara-based regression
# - Examine genotype-specific regression lines
```

- **Model Summary**: Displays the overall R-squared and residual statistics for
  the joint regression model using both the environmental mean and kPara as the
  environmental index.

- **Regression Lines**: Individual genotype regression lines overlaid on the
  observed data. The slope and intercept for each genotype are reported in a
  companion table.

### 6. CV Tab

The CV (Cross-Validation) tab evaluates prediction robustness:

```{r cv-example}
# In the CV tab:
# - Run environment leave-one-out cross-validation (cv_env)
# - Run observation-level LOOCV (loocv)
# - Compare prediction accuracy metrics
```

- **Environment CV**: Holds out one environment at a time and predicts its
  trait values from the remaining environments. Reports the correlation between
  observed and predicted values.

- **LOOCV**: Holds out one observation at a time for a finer-grained
  assessment. This is more computationally intensive but provides a
  per-observation prediction accuracy estimate.

- **Results Table**: Side-by-side comparison of accuracy metrics from both
  cross-validation approaches.

### 7. JGRA Tab

The JGRA (Joint Genomic Regression Analysis) tab integrates genotype marker
data with the CERIS environmental characterization. This tab is only available
when genotype data is present.

```{r jgra-example}
# In the JGRA tab:
# - Select method: RM.E (env LOO), RM.G (genotype CV), RM.GE (combined)
# - Choose between reaction norm (jgra) and marker effect (jgra_marker) models
# - Set fold and reshuffle parameters
# - View prediction accuracy and observed-vs-predicted plots
# - Run forecast_next_year with custom train/test splits
```

- **Method Selection**: Choose the cross-validation scheme. `RM.E` evaluates
  environment-level prediction, `RM.G` evaluates genotype-level prediction, and
  `RM.GE` evaluates both simultaneously.

- **Model Type**: Toggle between the reaction norm approach (`jgra`) and the
  marker effect approach (`jgra_marker`).

- **Fold and Reshuffle**: Set the number of cross-validation folds and
  reshuffles. Higher values give more stable estimates but increase computation
  time.

- **Forecast**: Define training and test environment sets for
  `forecast_next_year()`. Drag environments between the training and test
  panels, then run the forecast to see predicted versus observed values.

- **Results**: Displays `r_within`, `r_across`, and a scatter plot of observed
  versus predicted values colored by environment.

### 8. Visualization Tab

The Visualization tab provides tools for exporting and customizing plots:

```{r viz-example}
# In the Visualization tab:
# - Select any plot generated in previous tabs
# - Customize colors, labels, and dimensions
# - Export as PNG, PDF, or SVG
# - Download result tables as CSV
```

- **Plot Gallery**: Thumbnails of all plots generated during the session. Click
  to enlarge and customize.

- **Export Options**: Choose output format (PNG, PDF, SVG), resolution, and
  dimensions. Click "Download" to save.

- **Table Export**: Download any results table (CERIS search results,
  slope-intercept values, cross-validation metrics, JGRA predictions) as a CSV
  file for further analysis in other software.

---

## Custom Data Format Requirements

When uploading your own data, each CSV file must follow a specific column
structure. The app validates all files on upload and reports any issues.

### Traits File

The traits file contains phenotypic observations at the plot level.

| Column | Type | Required | Description |
|---|---|---|---|
| `line_code` | character | Yes | Genotype identifier |
| `env_code` | character | Yes | Environment identifier |
| *trait columns* | numeric | Yes (at least one) | One or more trait measurements (e.g., `FT`, `PH`, `GY`) |
| `pop_code` | character | No | Population identifier (optional grouping variable) |
| `env_note` | character | No | Short description of the environment |

Example:

```{r traits-format}
# line_code, env_code, FTdap, FTgdd, env_note
# "LINE001", "ENV01", 65.2, 850.1, "Lubbock 2018"
# "LINE001", "ENV02", 72.4, 920.3, "College Station 2018"
# "LINE002", "ENV01", 63.8, 835.6, "Lubbock 2018"
```

### Environment Metadata File

The environment metadata file describes each trial environment.

| Column | Type | Required | Description |
|---|---|---|---|
| `env_code` | character | Yes | Environment identifier (must match traits file) |
| `lat` | numeric | Yes | Latitude of trial site |
| `lon` | numeric | Yes | Longitude of trial site |
| `PlantingDate` | character/Date | Yes | Planting date (YYYY-MM-DD format) |
| `Year` | integer | No | Trial year |
| `Location` | character | No | Location name |

Example:

```{r env-meta-format}
# env_code, lat, lon, PlantingDate, Year, Location
# "ENV01", 33.57, -101.85, "2018-05-15", 2018, "Lubbock"
# "ENV02", 30.63, -96.33, "2018-04-20", 2018, "College Station"
```

### Environmental Parameters File

The environmental parameters file contains daily environmental covariates for
each environment.

| Column | Type | Required | Description |
|---|---|---|---|
| `env_code` | character | Yes | Environment identifier (must match traits file) |
| `DAP` | integer | Yes | Days after planting (1, 2, 3, ...) |
| *parameter columns* | numeric | Yes (at least one) | Daily environmental measurements |

Common parameter columns include:

| Parameter | Description |
|---|---|
| `TMAX` | Daily maximum temperature |
| `TMIN` | Daily minimum temperature |
| `DL` | Day length (photoperiod in hours) |
| `GDD` | Growing degree days |
| `PTT` | Photothermal time |
| `PTR` | Photothermal ratio |
| `PTS` | Photothermal sum |
| `PTD1` | Photothermal day formulation 1 |
| `PTD2` | Photothermal day formulation 2 |

The file must contain daily records for every environment, starting from DAP 1.
The maximum DAP should be at least as large as the `max_days` setting used in
the CERIS search.

Example:

```{r env-params-format}
# env_code, DAP, DL, GDD, PTT, PTR, PTS
# "ENV01", 1, 13.5, 15.2, 205.2, 1.13, 17.18
# "ENV01", 2, 13.5, 16.8, 226.8, 1.25, 21.00
# "ENV02", 1, 13.2, 18.1, 238.9, 1.37, 24.80
```

### Genotype File (Optional)

The genotype file contains marker scores for each genotype. If omitted, the
JGRA tab will be disabled.

| Column | Type | Required | Description |
|---|---|---|---|
| `line_code` | character | Yes | Genotype identifier (must match traits file) |
| *marker columns* | numeric | Yes | One column per marker, values typically coded as -1, 0, 1 |

Row names should correspond to `line_code` values in the traits file. The
marker values are typically coded as:

- `-1`: Homozygous for the reference allele
- `0`: Heterozygous
- `1`: Homozygous for the alternative allele

Example:

```{r genotype-format}
# line_code, MKR001, MKR002, MKR003, ...
# "LINE001", -1, 0, 1, ...
# "LINE002", 1, 1, -1, ...
# "LINE003", 0, -1, 0, ...
```

### Data Validation Checklist

Before uploading, verify that your data meets these requirements:

1. All `env_code` values in the traits file must appear in both the environment
   metadata and environmental parameters files.
2. All `line_code` values must be consistent across the traits and genotype
   files (if genotype data is provided).
3. The environmental parameters file must have continuous DAP values starting
   from 1 for each environment.
4. No missing values are allowed in the `env_code`, `line_code`, or `DAP`
   columns. Missing trait or parameter values should be coded as `NA`.
5. Column names must match exactly (case-sensitive).

You can also validate your data programmatically before launching the app:

```{r validate}
validate_input_data(
  traits     = my_traits,
  env_meta   = my_env_meta,
  env_params = my_env_params,
  genotype   = my_genotype  # omit if not available
)
```

## Tips for Effective Use

- **Start with a built-in dataset** to familiarize yourself with the workflow
  before uploading custom data.
- **Use the Explore tab** before running the CERIS search. Understanding the
  structure of your data helps interpret the search results.
- **Set max_days conservatively** (e.g., 80) for initial exploration. You can
  increase it later if the best window falls near the boundary.
- **Enable leave-one-out** in the CERIS tab for a robust assessment, but be
  prepared for longer computation times.
- **Export results** from the Visualization tab for publication-quality figures
  and for further analysis in R or other software.

## Summary

The CERIS Shiny application provides an interactive interface for the complete
analysis pipeline:

1. **Data**: Load built-in or custom datasets.
2. **Explore**: Visualize geographic patterns, trait distributions, and
   environmental profiles.
3. **CERIS**: Run the exhaustive window search and identify the critical
   environmental factor.
4. **Reaction Norm**: Examine genotype-specific environmental responses.
5. **JRA**: Fit joint regression models.
6. **CV**: Evaluate prediction accuracy through cross-validation.
7. **JGRA**: Perform genomic prediction (when genotype data is available).
8. **Visualization**: Export plots and result tables.

For reproducible analyses and scripting, all app functionality is available
through the CERIS R functions documented in the package vignettes.
